---
title: "Analysis of Wildlife Strikes to Aircraft"
author: "Vanshika Jain"
date: "Spring 2024"
output:
  pdf_document: default
  html_notebook: default
  html_document:
    df_print: paged
email: jain.van@northeastern.edu
subtitle: Practicum I CS5200
---
## Installing all required packages
```{r installer,echo = FALSE, eval=TRUE,warning=FALSE}
packages <- c("DBI", "tidyverse", "sqldf", "dplyr", "hash", "stringr", "RMySQL", "ggplot2", "RSQLite", "lubridate", "tidyr","tinytex")
# Install packages not yet installed
installed_packages <- packages %in% rownames(installed.packages())
if (any(installed_packages == FALSE)) {
  install.packages(packages[!installed_packages])
}

# Packages loading
invisible(lapply(packages, library, character.only = TRUE))
options(sqldf.driver = "SQLite")

```
## connecting to the cloud database hosted on gcloud
```{r conn,echo = FALSE, eval=TRUE,warning=FALSE}
# Create a connection to the SQLite database 
host = ""
user = "root"
port <- 3306

condb <- dbConnect(MySQL(), 
                 host = host, 
                 port= port,
                 user = user, 
                 dbname = "strikes")

```
## set constraint FK and drop tables
```{r drop,echo = FALSE, eval=TRUE,warning=FALSE}
# turn off foreign key constraint to drop tables
dbExecute(condb,"SET foreign_key_checks = 0;")

dbExecute(condb, "DROP TABLE IF EXISTS airports")
dbExecute(condb, "DROP TABLE IF EXISTS flights")
dbExecute(condb, "DROP TABLE IF EXISTS strikes")
dbExecute(condb, "DROP TABLE IF EXISTS conditions")
```
## create new tables
```{r create,echo = FALSE, eval=TRUE,warning=FALSE}
# create all tables - first create airports table for references
# turn on foreign key constraint to create connections
dbExecute(condb,"SET foreign_key_checks = 1;")
# airports table
dbExecute(condb, "CREATE TABLE IF NOT EXISTS airports(
                    aid INT PRIMARY KEY NOT NULL,
                    airportName TEXT,
                    airportState TEXT,
                    airportCode VARCHAR(3) DEFAULT 'ZZZ');
")

# flights table
dbExecute(condb, "CREATE TABLE IF NOT EXISTS flights (
                  fid INT PRIMARY KEY NOT NULL,
                  month INT NOT NULL,
                  origin INT NOT NULL,
                  airlineName TEXT,
                  aircraftType TEXT,
                  isHeavy BOOLEAN NOT NULL,
                  CONSTRAINT origin_fk FOREIGN KEY (origin)  REFERENCES airports(aid)
);
")
# conditions lookup table
dbExecute(condb, "CREATE TABLE IF NOT EXISTS conditions (
                  cid INT AUTO_INCREMENT PRIMARY KEY,
                  sky_condition VARCHAR(255) DEFAULT 'Unknown',
                  explanation VARCHAR(255) DEFAULT 'Unknown'
);
")
# strikes table
dbExecute(condb, "CREATE TABLE IF NOT EXISTS strikes (
                  sid INT AUTO_INCREMENT PRIMARY KEY,
                  fid INT NOT NULL,
                  numbirds INT DEFAULT 0,
                  impact VARCHAR(255) DEFAULT 'None',
                  damage BOOLEAN NOT NULL,
                  altitude INT CHECK (altitude >= 0) NOT NULL,
                  conditions INT NOT NULL,
                  CONSTRAINT flight_id FOREIGN KEY (fid) REFERENCES flights(fid),
                  CONSTRAINT sky_id FOREIGN KEY (conditions) REFERENCES conditions(cid)
);
")
```
## check the table info
```{r check,echo = FALSE, eval=TRUE,warning=FALSE}
# check the table info
air <- dbGetQuery(condb, "DESCRIBE airports")
fl <- dbGetQuery(condb, "DESCRIBE flights")
sky <- dbGetQuery(condb, "DESCRIBE conditions")
str <- dbGetQuery(condb, "DESCRIBE strikes")

checkAir <- dbGetQuery(condb, "SELECT * FROM airports LIMIT 10")
checkflt <- dbGetQuery(condb, "SELECT  * FROM flights LIMIT 10")
checkstr <- dbGetQuery(condb, "SELECT * FROM strikes LIMIT 10")
checksky <- dbGetQuery(condb, "SELECT * FROM conditions")

```
## read the csv data file
```{r readfile,echo = FALSE, eval=TRUE,warning=FALSE}
# read the csv data file and remove blanks or null rows
bds.raw <- read.csv(file = "BirdStrikesData-V3.csv", header = TRUE, stringsAsFactors = FALSE, strip.white = TRUE, sep=',')

bds.raw <- bds.raw[!(is.na(bds.raw$model) | bds.raw$model == ""), ]
bds.raw <- bds.raw[!(is.na(bds.raw$aircraft) | bds.raw$aircraft == ""), ]
bds.raw <- bds.raw[!(is.na(bds.raw$airline) | bds.raw$airline == ""), ]

df <- bds.raw[c("rid", "flight_date", "aircraft", "airline","airport", "impact", "damage", "origin", "sky_conditions", "altitude_ft", "heavy_flag")]

print(names(df))
```
## find the unique sky conditions to add in the table
```{r sky,echo = FALSE, eval=TRUE,warning=FALSE}
# Insert data into the conditions table
sky_conditions <- unique(df[c("sky_conditions")])
print(sky_conditions)
```
## insert data into conditions table
```{r condition,echo = FALSE, eval=TRUE,warning=FALSE}
dbExecute(condb, "INSERT INTO conditions (cid, sky_condition) VALUES (1,'No Cloud'), (2,'Some Cloud'), (3,'Overcast');")
# check the updated table
checksky <- dbGetQuery(condb, "SELECT * FROM conditions")
```
## insert data into airports 
```{r airports,echo = FALSE, eval=TRUE,warning=FALSE}
# add data into airports with some functions to handle extra strings 
data = unique(df[c("airport", "origin")])
data$aid = rownames(data)

airports <- data.frame(airportName = data$airport, airportState = data$origin, aid = data$aid)

airports$airportName <- sapply(airports[,'airportName'], function(x){if (grepl("\\'",x)){return(gsub("'", "", x, fixed=TRUE))} else {return(x)} })

airports$airportState <- sapply(airports[,'airportState'], function(x){if(grepl("\\'",x)) {return(gsub("'", "", x, fixed=TRUE))} else {return(x)}  })

# write this data frame into the table airports
dbWriteTable(condb, "airports", airports,row.names=FALSE, append=TRUE)

checkAir <- dbGetQuery(condb, "SELECT DISTINCT * FROM airports LIMIT 10")

```
## insert data into flights table
```{r temp,echo = FALSE, eval=TRUE,warning=FALSE}
# creating a temp table to store rid with origin and airport name
temp = data.frame(unique(df[c("rid", "airport", "origin")]))
dbExecute(condb, "DROP TABLE IF EXISTS x")
dbExecute(condb, "CREATE TABLE x(
                  rid INT,
                  origin VARCHAR(255),
                  airport VARCHAR(255));")

dbWriteTable(condb, "x", temp, row.names=FALSE, append = TRUE)
```
```{r tempflight,echo = FALSE, eval=TRUE,warning=FALSE}
# add data into flights table with aid FK
dbExecute(condb, "SET foreign_key_checks = 1;")

flights = unique(df[c("rid","flight_date", "heavy_flag", "airline", "aircraft")])

# creating a df with proper col names 
dfb = data.frame(fid = flights$rid, flight_date = flights$flight_date ,airlineName = flights$airline, aircraftType = flights$aircraft, isHeavy = flights$heavy_flag )

# changing the is heavy col with bool values
dfb$isHeavy = sapply(flights[, 'heavy_flag'], function(x){
  if(x=='Yes'){return(TRUE)} else{return(FALSE)}
})
dfb$isHeavy = as.integer(dfb$isHeavy)

# find the month and append a col for later use
dfb$month = sapply(flights[, 'flight_date'], function(x){
  val <- strsplit(x," ")[[1]][1]
  month <- strsplit(val,"/")[[1]][1]
  date <- strsplit(val,"/")[[1]][2]
  year <- strsplit(val,"/")[[1]][3]
  output = paste(year,"-",month,"-",date,sep="")
  return(month(ymd(output))) 
  })
```

```{r flights,echo = FALSE, eval=TRUE,warning=FALSE}
# creating another dummy table to store the imp info needed for flights and join with previous table
dbExecute(condb, "DROP TABLE IF EXISTS y")
dbExecute(condb, "CREATE TABLE y(
                  fid INT,
                  flight_date DATE,
                  airlineName TEXT,
                  aircraftType TEXT,
                  isHeavy BOOL,
                  month INT);")


dbWriteTable(condb, "y", dfb, row.names=FALSE, append = TRUE)

# getting all the cols from both tables using inner join on rid and airport name and state
final <- dbGetQuery(condb, "select * from 

                    (select a.rid,b.aid from airports b
                           INNER JOIN x a 
                           ON a.airport = b.airportName and a.origin = b.airportState) a
                           
                    inner join y b
                    on a.rid = b.fid")

# drop the extra tables
dbExecute(condb, "DROP TABLE IF EXISTS x")
dbExecute(condb, "DROP TABLE IF EXISTS y")

# just select the required cols and remove redundant info
final = final[c("fid", "aid", "airlineName", "aircraftType", "isHeavy", "month")]

# renamed the col as required
names(final)[names(final) == "aid"] <- "origin"

dbWriteTable(condb, "flights", final, row.names=FALSE, append = TRUE)

checkflt <- dbGetQuery(condb, "SELECT * FROM flights LIMIT 10")
```

## insert data into strikes table
```{r strikes,echo = FALSE, eval=TRUE,warning=FALSE}
# add data into strikes
strikes =  unique(bds.raw[c("rid","impact", "damage", "altitude_ft")])

# creating a df with proper col names needed
striketable = data.frame(fid = strikes$rid, sid=rownames(strikes), impact = strikes$impact, damage = strikes$damage, altitude = strikes$altitude_ft)

# remove extra commas
striketable$altitude = sapply(striketable[, 'altitude'], function(x){if(grepl("\\,",x)){return(gsub(",", "", x, fixed = TRUE))} else{return(x)}
  })

# change it to boolean
striketable$damage = sapply(striketable[,'damage'], function(x){ if(x=='No damage'){return(FALSE)} else{return(TRUE)}
  })
striketable$damage = as.integer(striketable$damage)

# create dummy table with all this info to join on another table with conditions cid
dbExecute(condb, "DROP TABLE IF EXISTS dummy")
dbExecute(condb,"create table dummy(
                  fid int,
                  sid int,
                  impact TEXT,
                  damage bool,
                  altitude int);")

dbWriteTable(condb, "dummy", striketable, row.names=FALSE, append=TRUE)

checkstr <- dbGetQuery(condb, "SELECT * FROM dummy LIMIT 10")
```

```{r strikestable,echo = FALSE, eval=TRUE,warning=FALSE}
# create a temp table with cid from conditions table
temp = data.frame(unique(df[c("rid", "sky_conditions")]))

dbExecute(condb, "drop table if exists sky")
dbExecute(condb, "create table sky(
                  rid int,
                  sky_conditions text);")

dbWriteTable(condb, "sky", temp, row.names=FALSE, append = TRUE)

sky = dbGetQuery(condb, "select * from sky")

# do a inner join on condition and store it in the final table strikes
skydata = dbGetQuery(condb,"SELECT * FROM dummy a
                            INNER JOIN 
                                  (SELECT b.cid,a.rid FROM sky a
                                  INNER JOIN conditions b
                                  ON a.sky_conditions=b.sky_condition) 
                                  b
                            ON a.fid = b.rid;
                            ")
dbExecute(condb, "drop table if exists sky")
dbExecute(condb, "drop table if exists dummy")

# select the needed cols
skydata = skydata[c("fid", "cid", "impact", "damage", "altitude", "sid")]
names(skydata)[names(skydata) == "cid"] = "conditions"

dbWriteTable(condb, "strikes", skydata, row.names=FALSE, append = TRUE)

checkstr = dbGetQuery(condb, "select * from strikes")

```

## Display tables
```{r dispairport,echo = FALSE,warning=FALSE, eval=TRUE}
print("Airport table")
print(head(dbGetQuery(condb, "SELECT * FROM airports LIMIT 10"), 5))
```

```{r dispflight,echo = FALSE,warning=FALSE, eval=TRUE}
print("Flight table")
print(head(dbGetQuery(condb, "SELECT * FROM flights LIMIT 10"), 5))
```

```{r dispsky,echo = FALSE,warning=FALSE, eval=TRUE}
print("conditions table")
print(head(dbGetQuery(condb, "SELECT * FROM conditions"), 5))

```

```{r dispstrike,echo = FALSE,warning=FALSE,eval=TRUE}
print("strikes table")
print(head(dbGetQuery(condb, "SELECT * FROM strikes LIMIT 10"), 5))

```
## Top Airports with Strikes
```{r top,echo = FALSE,warning=FALSE,eval=TRUE}
# top 10 airports are found using two inner joins on flights and airports to group by state
top10Airports <- dbGetQuery(condb, "SELECT a.airportState AS state, COUNT(s.sid) AS number_of_incidents
                                    FROM strikes s
                                    INNER JOIN flights f ON s.fid = f.fid
                                    INNER JOIN airports a ON f.origin = a.aid
                                    GROUP BY a.airportState
                                    ORDER BY number_of_incidents DESC, a.airportState ASC
                                    LIMIT 10;
                                    
                                    
")
print((top10Airports))
```

## Analysis by Airline
```{r analysisAir,echo = FALSE,warning=FALSE,eval=TRUE}
# airlines with more than average num of strikes by inner join on flights and grouping by airlines
strikePerAirline <- dbGetQuery(condb,"SELECT f.airlineName AS airline, 
                                      COUNT(s.sid) AS num_incidents
                                      FROM strikes s
                                      INNER JOIN flights f ON s.fid = f.fid
                                      GROUP BY f.airlineName
                                      HAVING COUNT(s.sid) > (SELECT AVG(num_incidents)
                                                            FROM (
                                                                SELECT COUNT(sid) AS num_incidents
                                                                FROM strikes
                                                                GROUP BY fid) AS avg_count);

                                      ")
print(strikePerAirline)
```

## Analysis by Month
```{r analmonth,echo = FALSE,warning=FALSE,eval=TRUE}
# strikes by month is just counting ids given the case of that month
monthlyStrikes <- dbGetQuery(condb, "
    SELECT 
        CASE 
            WHEN f.month = 1 THEN 'January'
            WHEN f.month = 2 THEN 'February'
            WHEN f.month = 3 THEN 'March'
            WHEN f.month = 4 THEN 'April'
            WHEN f.month = 5 THEN 'May'
            WHEN f.month = 6 THEN 'June'
            WHEN f.month = 7 THEN 'July'
            WHEN f.month = 8 THEN 'August'
            WHEN f.month = 9 THEN 'September'
            WHEN f.month = 10 THEN 'October'
            WHEN f.month = 11 THEN 'November'
            WHEN f.month = 12 THEN 'December'
        END AS month,
        COUNT(f.fid) AS num_strikes
    FROM flights f
    GROUP BY f.month
    ORDER BY f.month
")

print(monthlyStrikes)

```
## Trend by Month
```{r trend,echo = FALSE,warning=FALSE,eval=TRUE}
# Aggregate the number of strikes by month
birds_by_month <- aggregate(monthlyStrikes$num_strikes, by = list(month = monthlyStrikes$month), FUN = sum)

# Create a horizontal bar chart
barplot(birds_by_month$x,
        names.arg = birds_by_month$month,
        horiz = TRUE,
        xlab = "Number of Birds",
        ylab = "Month",
        main = "Number of Birds by Month",
        col = "skyblue",
        border = "black",
        las = 1,  # Rotate axis labels
        xlim = c(0, max(birds_by_month$x) * 1.1),  # Adjust x-axis limits
        cex.names = 0.8,  # Reduce size of month labels
        cex.axis = 0.8,  # Reduce size of axis labels
        cex.lab = 1.4  # Increase size of axis labels
)

# Add data labels
text(birds_by_month$x, 1:length(birds_by_month$month), labels = birds_by_month$x, pos = 4)

# Add legend
legend("bottomright", legend = "Number of Birds", fill = "skyblue", bty = "n")

```
## Test the stored procedure by removing a strike

```{r procedure, echo=FALSE,eval=TRUE}
dbExecute(condb, "DROP PROCEDURE if exists RemoveStrike;")
dbExecute(condb, "CREATE PROCEDURE RemoveStrike(IN strike_id INT)
BEGIN
    DELETE FROM strikes WHERE sid = strike_id;
    INSERT INTO audit_log (modification_type, table_name, modification_time, primary_key_str)
    VALUES ('Removal', 'strikes', CURRENT_TIMESTAMP, strike_id);
END ")
```

```{r delete, echo=FALSE,eval=TRUE}
# Create the audit log table if it does not exist
dbExecute(condb,"CREATE TABLE IF NOT EXISTS audit_log (
    id INT AUTO_INCREMENT PRIMARY KEY,
    modification_type VARCHAR(20),
    table_name VARCHAR(50),
    modification_time DATETIME,
    primary_key_str INT
);"
)

dbExecute(condb, "CALL RemoveStrike(2)")
# Query the audit log to verify the removal is logged
audit_log <- dbGetQuery(condb, "SELECT * FROM audit_log")
print(audit_log)


```

```{r exit,echo=FALSE}

# Close the database connection when done
dbDisconnect(condb)
```

